---
title: "Simulation-Based Calibration for an upsis Regression Update"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation-Based Calibration for an upsis Regression Update}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  eval = FALSE
)
```

This article sketches a simulation-based calibration (SBC) workflow for the
core `upsis` use case: fit a simple `brms` regression to an initial data batch,
then use PSIS to incorporate a second data batch without refitting.

The example uses 100 observations per simulated dataset. Each fit sees the first
50 observations directly, and `upsis()` incorporates observations 51 through
100. All priors are proper and explicitly declared in both the simulator and the
`brms` model:

- intercept: `normal(0, 2)`;
- slope: `normal(0, 1)`;
- residual scale: `exponential(1)`.

The chunks are not evaluated when the package is checked, because a useful SBC
run requires many repeated Stan fits. Copy them into an interactive session or
temporarily set `eval = TRUE` while rendering locally. During development,
reduce `n_sims` before rendering the full article.

```{r setup}
library(SBC)
library(brms)
library(ggplot2)
library(posterior)
library(upsis)

options(brms.backend = "cmdstanr")
```

```{r backend-result}
new_upsis_sbc_fit <- function(draws, pareto_k) {
  structure(
    list(draws = draws, pareto_k = pareto_k),
    class = "upsis_sbc_fit"
  )
}

as_draws_matrix.upsis_sbc_fit <- function(x, ...) {
  x$draws
}
```

```{r priors}
regression_priors <- prior(normal(0, 2), class = "Intercept") +
  prior(normal(0, 1), class = "b", coef = "x") +
  prior(exponential(1), class = "sigma")
```

## Data generator

The generator draws parameters from the declared priors, simulates 100 predictor
values, and then simulates the response from the linear Gaussian model.

```{r generator}
generate_regression_data <- function(n = 100) {
  b_Intercept <- rnorm(1, mean = 0, sd = 2)
  b_x <- rnorm(1, mean = 0, sd = 1)
  sigma <- rexp(1, rate = 1)

  x <- rnorm(n)
  y <- rnorm(n, mean = b_Intercept + b_x * x, sd = sigma)

  list(
    variables = list(
      b_Intercept = b_Intercept,
      b_x = b_x,
      sigma = sigma
    ),
    generated = data.frame(y = y, x = x)
  )
}

generator <- SBC_generator_function(generate_regression_data)
```

```{r datasets}
set.seed(20241109)
n_sims <- 50
datasets <- generate_datasets(generator, n_sims)
```

## Backend: fit first, update second

This backend is the algorithm under calibration. For each simulated dataset, it
fits the regression to the first 50 observations, then applies `upsis()` to add
the remaining 50 observations.

```{r backend}
fit_then_update <- function(generated, cores = 1) {
  initial_data <- generated[1:50, , drop = FALSE]
  update_data <- generated[51:100, , drop = FALSE]

  initial_fit <- brm(
    y ~ x,
    data = initial_data,
    prior = regression_priors,
    family = gaussian(),
    chains = 4,
    iter = 3500,
    warmup = 1000,
    cores = cores,
    seed = 123,
    refresh = 0
  )

  updated <- upsis(initial_fit, data_add = update_data)

  posterior_draws <- posterior::as_draws_matrix(
    updated$updated_model,
    variable = c("b_Intercept", "b_x", "sigma")
  )

  new_upsis_sbc_fit(
    draws = posterior_draws,
    pareto_k = updated$pareto_k
  )
}

backend <- SBC_backend_function(
  fit_then_update,
  cores_arg = "cores",
  default_thin_ranks = 10
)
```

## Run SBC

```{r compute}
results <- compute_SBC(
  datasets,
  backend,
  keep_fits = TRUE,
  cache_mode = "results",
  cache_location = file.path(tempdir(), "upsis-sbc-linear-regression")
)
```

The default R-hat and ESS diagnostics reported by `compute_SBC()` are computed
from the posterior draws returned by the backend. Here, those are the draws from
`updated$updated_model`, after incorporating observations 51 through 100 with
`upsis()`. They are not diagnostics from the initial 50-observation `brms` fit.

Because `keep_fits = TRUE` stores the compact objects returned by the backend,
we can also summarize the Pareto `k` values from the PSIS update:

```{r pareto-k-summary}
pareto_k <- vapply(
  results$fits,
  function(fit) {
    if (inherits(fit, "upsis_sbc_fit")) {
      fit$pareto_k
    } else {
      NA_real_
    }
  },
  numeric(1)
)

summary(pareto_k)
mean(pareto_k > 0.7, na.rm = TRUE)
```

```{r rank-histogram, fig.width = 10, fig.height = 4.8}
plot_rank_hist(results)
```

```{r ecdf-difference, fig.width = 10, fig.height = 4.8}
plot_ecdf_diff(results)
```

The rank histograms and ECDF difference plots should be read alongside the
ordinary fit diagnostics and the Pareto `k` warnings produced by `upsis()`. A
high rate of problematic Pareto `k` values means this particular update is too
large or too influential for importance sampling to replace a full refit.
